---
title: "TidyTuesday -- Powerlifting Performance & Age"
author: "Patrick Ward"
date: "1/14/2020"
output: html_document
---
---
title: "Powerlifting Tidy Tuesday"
author: "Patrick Ward"
date: "1/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(scales)

theme_set(theme_bw())
```

# Load data

```{r}
df <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-10-08/ipf_lifts.csv") %>%
  filter(age_class != "5-12", !is.na(age))

df %>% head(10) %>% View()

# how many rows of data
nrow(df)

# how many unique athletes
nrow(distinct(df, name))
```

# Data Clean Up

```{r}
# Some of the values for the lifts are negative, which is impossible. Turn those into 0's

df <- df %>%
  mutate(best3squat_kg = ifelse(best3squat_kg < 0, 0, best3squat_kg),
        best3bench_kg = ifelse(best3bench_kg < 0, 0, best3bench_kg),
        best3deadlift_kg = ifelse(best3deadlift_kg < 0, 0, best3deadlift_kg))

# order the data by lifter and date and create a total of their weight lifting in a meet

df <- df %>%
  mutate(Total = best3squat_kg + best3bench_kg + best3deadlift_kg) %>%
  arrange(name, date)

# create a long format of the data for plotting purposes

df_long <- df %>%
  reshape2::melt(., id = c("name", "date", "age", "age_class", "weight_class_kg", "sex"), measure.vars = c("best3squat_kg", "best3bench_kg", "best3deadlift_kg", "Total")) %>%
  na.omit(df_long)

```

# Data Exploration

```{r}
# how many competitions have athletes competed in?
df %>%
  count(name, sort = T) %>%
  head(5) %>%
  mutate(name = fct_reorder(name, n)) %>%
  ggplot(aes(x = name, y = n)) +
  geom_col(aes(fill = n)) +
  coord_flip() +
  scale_fill_gradientn(colors = rainbow(2))

# Most common type of equipment
df %>%
  count(equipment, sort = T) %>%
  mutate(equipment = fct_reorder(equipment, n)) %>%
  ggplot(aes(x = equipment, y = n)) +
  geom_col() +
  coord_flip()

## what age does the best lift happen at?
# Age Class
df_long %>%
  ggplot(aes(x = age_class, y = value)) +
  geom_boxplot() +
  facet_wrap(~variable, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

# Age (numeric)
df_long %>%
  ggplot(aes(x = age, y = value)) +
  geom_point(color = "grey", alpha = 0.7) +
  geom_smooth(se = F) +
  facet_wrap(~variable, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

# Age (numeric) by sex
df_long %>%
  ggplot(aes(x = age, y = value)) +
  geom_point(color = "grey", alpha = 0.7) +
  geom_smooth(se = F) +
  facet_wrap(sex~variable, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

# Distribution of Totals
df %>%
  ggplot(aes(x = Total)) +
  geom_density(aes(fill = sex), alpha = 0.3) +
  labs(title = "Distribution of Powerlifting Totals")


```


# Research questions

- We see some evidence that the athlete's peak around their early 30's
- Can we build a statistical model to evaluate the observed aging curve in powerlifters?
- Using a "time to event" analysis, can we identify how many competitions a lifter competes in until achiving an elite Total (the event of interest)?

# Data Clean Up Part 2

- Get only single-ply lifters
- Filter out only those athletes who have competed in 10 or more competitions


```{r}
sply <- df %>%
  filter(equipment == "Single-ply") %>%
  group_by(name) %>%
  filter(n() >= 10)

nrow(sply)
nrow(distinct(sply, name))

# 6169
# 426 athletes
```


# Models

##### Powerlifter Aging Curve

- NOTE: I did not account for sex in the analysis, just to keep things simple.
- This model can be extended to a mixed model in the future.

```{r}

## create a peak age function, that can calculate the peak agre from the coefficients of the polynmial model

peak_age <- function(coef1, coef2){
  x = -(coef1) / (2 * (coef2))
  return(x)
}

## Build a polynomial model for each lift
squat_age_fit <- lm(best3squat_kg ~ age + I(age^2), data = sply)
bench_age_fit <- lm(best3bench_kg ~ age + I(age^2), data = sply)
deadlift_age_fit <- lm(best3deadlift_kg ~ age + I(age^2), data = sply)

## Get model summaries
summary(squat_age_fit)
summary(bench_age_fit)
summary(deadlift_age_fit)

## Calculate peak ages with the peak_age() function
sq_peak_age <- peak_age(coef1 = squat_age_fit$coef[2], coef2 = squat_age_fit$coef[3])
bench_peak_age <- peak_age(coef1 = bench_age_fit$coef[2], coef2 = bench_age_fit$coef[3])
deadlift_peak_age <- peak_age(coef1 = deadlift_age_fit$coef[2], coef2 = deadlift_age_fit$coef[3])

sq_peak_age
bench_peak_age
deadlift_peak_age

## Plot an example of the bench model

age = seq(
  from = range(sply$age)[1],
  to = range(sply$age)[2],
  by = 1)

pred <- predict(bench_age_fit, newdata = data.frame(age))
CL_95 <- 1.96 * predict(bench_age_fit, newdata = data.frame(age), se = T)$se.fit

df_pred <- data.frame(age, pred, CL_95)

df_pred %>%
  ggplot(aes(x = age, y = pred)) +
  geom_point(aes(x = age, y = best3bench_kg), data = sply, color = "grey", alpha = 0.2) +
  geom_ribbon(aes(ymin = pred - CL_95, ymax = pred + CL_95),
              fill = "green", alpha = 0.5) +
  geom_line(color = "red", size = 1.2) +
  annotate("text", x = 35.5, y = 172.25507, label = "X", size = 12) +
  labs(x = "age", y = "bench press", title = "Model Predicted Bench Press Line vs Actual Outcomes (dots)")

```


##### Time to elite status

- Create a classification for elite

https://www.lift.net/2013/05/09/classification-standards-for-raw-elite-uspa/

```{r}

df_elite <- df %>%
  filter(sex == "M") %>%
  mutate(elite_total = 
    ifelse(weight_class_kg == "52", 475,
    ifelse(weight_class_kg == "56", 515,
    ifelse(weight_class_kg == "60", 555,
    ifelse(weight_class_kg == "67.5", 622,
    ifelse(weight_class_kg == "74" | weight_class_kg == "75", 681,
    ifelse(weight_class_kg == "82.5", 732,
    ifelse(weight_class_kg == "90" | weight_class_kg == "93", 772,
    ifelse(weight_class_kg == "100", 814,
    ifelse(weight_class_kg == "110" | weight_class_kg == "105", 843,
    ifelse(weight_class_kg == "125" | weight_class_kg == "120", 868,
    ifelse(weight_class_kg == "125+", 907, NA))))))))))))


# remove NA is elite_total column

df_elite <- df_elite %>% filter(!is.na(elite_total))

# Create an "is_elite" tag

df_elite <- df_elite %>%
  mutate(is_elite = ifelse(Total >= elite_total, 1, 0)) %>%
  arrange(name, date)

## Create a comp_id column to count each competition the lifter was in

df_elite <- df_elite %>%
  group_by(name) %>%
  mutate(comp_id = seq_along(name))

## create a data set that has whether the first time the lifter achieved an elite total or the total number of competitions the athlete has been in without achieving an elite total

df_elite2 <- df_elite %>%
  group_by(name) %>%
  filter(cumsum(is_elite) == 0 | (cumsum(is_elite) == 1 & is_elite == 1)) %>%
  filter(comp_id == max(comp_id)) %>%
  select(name, age, age_class, Total, elite_total, comp_id, is_elite)

nrow(distinct(df_elite2, name))

# 6074 unique lifters

table(df_elite2$is_elite)
prop.table(table(df_elite2$is_elite))
barplot(table(df_elite2$is_elite), main = "Number of Male Lifters Totaling Elite\n(0 = Non-Elite Total / 1 = Elite Total)")

df_elite2 %>%
  ggplot(aes(x = comp_id, y = is_elite)) +
  geom_bar(stat = "identity") +
  labs(x = "Competition Number",
       y = "Lifters Totaling Elite",
       title = "Number of Competitions until a Lifter Totals Elite")

## Time-to-event model

library(survival)
library(broom)

model_total <- survfit((Surv(comp_id, is_elite) ~ 1), data = df_elite2)

summary(model_total)

tidy(model_total) %>%
  ggplot(aes(x = time, y = estimate)) +
  geom_line() +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = .2) +
  scale_y_continuous(labels = percent_format()) +
  expand_limits(y = .10) +
  ggtitle("Time to Elite Total")

```


